---
title: "Genetic markers combined with air pollution exposure in the development of lung cancer in a Northern Israeli cohort of different ethnic origins"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

Introduction
============

Lung cancer (LC) remains the leading cause of cancer mortality worldwide and poses a major public health burden. The development of LC is driven by a combination of genetic susceptibility and environmental exposures, particularly air pollution, which is recognized as a Group 1 carcinogen by the International Agency for Research on Cancer (IARC). 

The **relevance** of this topic is multidimensional. Clinically, LC represents a critical challenge in oncology due to its heterogeneous nature and frequent resistance to targeted therapies. Epidemiologically, it highlights the interaction between environmental determinants and host factors in shaping disease risk.

Risk factors for LC in Israel align with global patterns but also show local nuances. ***Tobacco smoking*** remains the predominant cause, implicated in an estimated 85% of cases. However, urban and occupational ***air pollution*** also plays a significant role. Haifa and its surrounding bay area are the Israeli most industrialized and heavily polluted region. Several studies have consistently reported elevated cancer risks, including LC, among its local heterogeneous population.

LC management has become increasingly dependent on molecular characterization, with several biomarkers guiding prognosis and therapy. The key **LC genetic markers** are the follows: 

1) programmed death-ligand 1 (PD-L1); its expression is measured by ***tumor proportion score (TPS)*** - the percentage of viable tumor cells that express PD-L1 on their surface
2) mesenchymal–epithelial transition proto-oncogene (MET)
3) epidermal growth factor receptor (EGFR)
4) B-Raf proto-oncogene (BRAF)
5) Kirsten rat sarcoma (KRAS)
6) erythroblastic oncogene B2 (ERBB2) (human epidermal growth factor receptor 2 (HER2))
7) anaplastic lymphoma kinase (ALK)
8) c-ros oncogene 1 (ROS1) 
9) neurotrophic tyrosine receptor kinases (NTRK) genes (NTRK1, NTRK2, and NTRK3)
10) rearranged during transfection (RET)
11) mismatch repair (MMR) genes (MLH1, PMS2, MSH2, MSH6); their abnormal function leads to instability of short tandem DNA repeats (***microsatellites***) evaluated by the ***microsatellite instability (MSI)*** status of the tumor

From a toxicological and epidemiological perspective, several **air pollutants** have been most consistently linked with LC incidence and mortality: 

1) ***NOx*** - nitrogen oxides generally (including NO and NO₂); most epidemiological studies do not separate NO from NO₂, so the evidence is dominated by NO₂; however, NOx may capture additional sources (e.g. industrial emissions) and serve as an upstream precursors to secondary pollutants (ozone, nitrates, etc.)
2) carbon monoxide (CO)
3) sulfur dioxide (SO₂) 
4) ozone (O₃)
5) particulate matter with the particles diameter ≤ 10 µm (PM₁₀); represents coarse plus fine particles
6) particulate matter with the particles diameter ≤ 2.5 µm (PM₂.₅) (fine particles) 

The **aim** of this work is to examine the combined influence of genomic alterations and environmental exposures on LC development in Northern Israel:

1. Molecular profiling: to characterize the prevalence of key molecular alterations in LC cases, including PD-L1 expression (via immunohistochemistry (IHC)), driver mutations (EGFR, KRAS, BRAF, MET, ALK, ROS1, ERBB2, RET, NTRK1/2/3; via next-generation sequencing (NGS)), and mismatch repair (MMR) genes (MLH1, PMS2, MSH2, MSH6; via polymerase chain reaction (PCR)).
2. Demographic stratification: to categorize cases by ethnic origin and examine differences in clinical or molecular characteristics across groups.
3. Geographic mapping: to document the residential distribution of patients within the Israeli northern districts and evaluate spatial tendencies in case occurrence, with further comparing them with spatially and temporally resolved air pollution estimates.

Data management
===============

## Data sources

1. **Air quality data** were obtained from the official website of the Israeli Ministry of Environmental Protection: https://air.sviva.gov.il. The most recent report openly available (2023) includes nationwide air concentration maps for major pollutants of established epidemiological relevance.

2. **Clinical-laboratory data** were obtained from the *anonymized* individual electronic medical records (EMRs) of *94 lung cancer patients* managed at Lady Davis Carmel Medical Center in Haifa in 2024-2025 and whose lung biopsy and surgical material were diagnosed in the Department of Pathology. Twenty-one variables were chosen, covering demographics (age, sex, ethnic group, residential district, smoking history), histopathological diagnosis, and molecular biomarkers (PD-L1 expression, EGFR, KRAS, BRAF, MET, ALK, ROS1, NTRK, RET, ERBB2, MSI, TMB). 

## Data setup

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)

paths <- list(
  raw = "data",
  figures = "output",
  tables = "output"
)
```

```{r}
df_raw <- read_excel(file.path(paths$raw, "CarmelMC_LungCancer.xlsx"))|>
  clean_names()

glimpse(df_raw)
```

Data dictionary
===============

The clinical-laboratory variables are described according to a ***syntactical approach***, distinguishing only between numeric and character. Under this approach, the functional role of a variable (continuous, discrete count, binary flag, or categorical) is determined by its analytical use rather than by its storage type.

- id (numeric, nominal; identifier, not for analysis): a clinical case's anonymized number
- sex (numeric, binary): a patient's biological sex: 0 = male; 1 = female
- age (numeric, continuous): a patient's age (years) at the date of the lung biopsy or surgery
- ethnic (numeric, nominal): a patient's presumed ethnic origin (subjective approximation based on the patient's name/surname evaluation and locality; not officially confirmed): 0 = Jewish; 1 = Arab, not otherwise specified (NOS); 2 = Druze; 3 = Bedouin   
- locality (character, nominal): a patient's place of residence (city, town, village, etc.)
- latitude (numeric, continuous): a locality’s geographic latitude (decimal degrees)
- longitude (numeric, continuous): a locality’s geographic longitude (decimal degrees)
- loc_popul (numeric, continuous): a locality’s population size according to the Israeli Central Bureau of Statistics (2023) (number of peoples)
- district (numeric, binary): Israeli administrative district to which a patient's locality belongs: 0 = Haifa District; 1 = Northern District
- smoking (numeric, binary): a patient's smoking status: 0 = non-smoker (never smoked or quit > 10 years before the lung biopsy/surgery date); 1 = smoker (current smoker or quit ≤ 10 years before the lung biopsy/surgery date)
- bx_date (numeric, ordinal): date of the lung biopsy or surgery performed
- dx (numeric, nominal): pathohistological diagnosis of lung cancer: 0 = adenocarcinoma (ACA), NOS; 1 = mucinous ACA (MucACA); 2 = squamous cell carcinoma (SCC); 3 = neuroendocrine tumour (NET) (low-grade (typical carcinoid) or intermediate-grade (atypical carcinoid) neuroendocrine neoplasm); 4 = neuroendocrine carcinoma (NEC) (high-grade neuroendocrine neoplasm (small cell lung cancer or large cell neuroendocrine carcinoma))  
pdl1 (numeric, binary) - PD-L1 expression measured by TPS:
0 = weak (TPS = 1-49%) 
1 = strong (TPS ≥ 50%)
met (numeric, binary) - MET gene mutation status: 
0 = no mutation 
1 = mutation detected
egfr (numeric, binary) - EGFR gene mutation status: 
0 = no mutation 
1 = mutation detected
braf (numeric, binary) - BRAF gene mutation status: 
0 = no mutation 
1 = mutation detected
kras (numeric, binary) - KRAS gene mutation status: 
0 = no mutation 
1 = mutation detected
erbb2 (numeric, binary) - ERBB2 (HER2) gene mutation status: 
0 = no mutation 
1 = mutation detected
alk (numeric, binary) - ALK gene mutation status: 
0 = no mutation 
1 = mutation detected
ros1 (numeric, binary) - ROS1 gene mutation status: 
0 = no mutation 
1 = mutation detected
ntrk (numeric, binary) - NTRK1/NTRK2/NTRK3 genes mutation status: 
0 = no mutation 
1 = mutation detected
ret (numeric, binary) - RET gene mutation status: 
0 = no mutation 
1 = mutation detected
msi (numeric, ordinal) - microsatellite instability (MSI) status of the tumour - instability of short tandem DNA repeats (microsatellites) of tumour cells, which results from abnormal function of one or more mismatch repair (MMR) genes (MLH1, PMS2, MSH2, and MSH6): 
0 = microsatellite stable (MSS) status 
1 = low microsatellite instability (MSI-L) 
2 = high microsatellite instability (MSI-H)
tmb (numeric, continuous) - tumour mutational burden (TMB) - number of non-inherited mutations per million bases (megabase, Mb) of the investigated genomic sequence. TMB is strongly predictive of a patient's response to immune checkpoint inhibitor therapy, of overall as well as disease-specific survival - independently of cancer type, stage or grade.








This step makes all coded variables human-readable and ensures models/tables treat them as categorical rather than numeric.

```{r}
library(dplyr)
library(forcats)

df <- df_raw |>
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("male", "female")),
    smoking = factor(smoking, levels = c(0, 1), labels = c("never smoked or quit >10y", "current smoker or quit ≤10y")),
    district = factor(district, levels = c(0, 1), labels = c("Haifa District", "Northern District")),
    pdl1     = factor(pdl1, levels = c(0, 1), labels = c("weak (TPS 1–49%)", "strong (TPS ≥50%)")),

    met  = factor(met,  levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    egfr = factor(egfr, levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    braf = factor(braf, levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    kras = factor(kras, levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    erbb2= factor(erbb2,levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    alk  = factor(alk,  levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    ros1 = factor(ros1, levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    ntrk = factor(ntrk, levels = c(0, 1), labels = c("no mutation", "mutation detected")),
    ret  = factor(ret,  levels = c(0, 1), labels = c("no mutation", "mutation detected")),

    msi = factor(msi, levels = c(0, 1, 2),
                 labels = c("MSS", "MSI-L", "MSI-H"),
                 ordered = TRUE),

    ethnic = factor(ethnic, levels = c(0, 1, 2, 3),
                    labels = c("Jewish", "Arab NOS", "Arab (Druze)", "Arab (Bedouin)")),

    dx = factor(dx, levels = c(0, 1, 2, 3, 4),
                labels = c("ACA, NOS", "mucinous ACA (MucACA)", "SCC",
                           "NET (low/intermediate-grade)", "NEC"))
  )

glimpse(df)
```
